### 2016-05-05 Thesis/DCG_sb-wi-thesis-template/MSpC/MSpC.tex
added tutorial.pdf

The Morpheus MS search engine was recently designed for high-resolution, accurate-mass data obtained from Orbitrap-based instruments to provide faster matching of spectra to peptides \citep{wenger13}.
Unfortunately, no downstream automated tools are available to facilitate LFQ analysis, which can be quite challenging, if not impossible, to complete manually when accounting for shared peptides.
To overcome this bottleneck, we developed Morpheus Spectral Counter (MSpC) as the first LFQ computational tool that integrates directly with Morpheus to calculate NSAF, dNSAF, uNSAF, and corrected PSM \citep{fermin11} values in complex protein samples.
\begin{sidewaysfigure}[p]
	\centering
	\includegraphics[width=\columnwidth]{MSpC/figure1_supplemental.png}
	\mycaption{MSpC Graphic User Interface (GUI) and software flow chart}{
\textbf(A) Screenshot of the GUI.
The input requires the user to select a Morpheus search summary file containing experiments to be analyzed.  The user can optionally select a whitelist to filter output, and select an output directory.
Additional options can set peptide and protein FDR cutoffs, and method of quantification for output, including Normalized Spectral Abundance Factor (NSAF), distributed NSAF (dNSAF), and unique	NSAF (uNSAF).
A progress bar highlights completion of the analysis.
\textbf(B) Data analysis flow chart.
Experiments and groups of experiments to be analyzed are imported through the Morpheus summary file.
PSMs and protein groups are filtered at the specified FDR cutoff with a default of 1\%.}
	\label{fig:GUIflow}
\end{sidewaysfigure}
MSpC is fully automated, and only requires a Morpheus search summary file (summary.tsv) as input.
The user interface (see Figure \ref{fig:GUIflow} (A)) allows one to select the summary file and displays the raw MS/MS files that will be analyzed by MSpC.
Due to shared peptides being attributed to only one instance of a protein group in Morpheus's PSM file, PSMs are re-matched to all possible protein groups.
PSMs are then cataloged as shared or as unique (distinctly matching one protein group) to generate NSAF, dNSAF, and uNSAF outputs.
Finally, the output can be filtered for proteins of interest by specifying a comma delimited file containing unique identifiers and descriptions.
Some important features of MSpC are its ability to handle fractionation experiments as input, and the ability to whitelist proteins of interest in the output by specifying a csv file (see Tutorial).
Options exist to specify global PSM and protein group FDR rates (thus avoiding increased FDRs when one analyzes many experiments at once), to output NSAF, dNSAF, and uNSAF values, to require a minimum number of unique peptides to quantify a protein, and to specify an output directory.
A progress bar indicates completion of the analysis by MSpC.
To validate the accuracy of MSpC, we analyzed two MS/MS datasets available in PRIDE that were previously generated by high-energy collision-induced dissociation using Thermo Q-Exactive Orbitrap instruments.
\begin{figure}[p]
	\centering
	\includegraphics[width=\columnwidth]{MSpC/figure1.png}
	\mycaption{Confirmation of MSpC accuracy by analysis of MS/MS datasets generated with the Universal Proteome Standard 2 (UPS2)}{The array of UPS2 standards were spiked into Xenopus laevis egg \textbf(Top) and embryo \textbf(Bottom) extracts at a range of concentrations.  Following MS/MS analysis, dNSAF values for each protein were determined by Morpheus and MSpC.  \textbf(A) A log-log plot of dNSAF versus concentration for each UPS2 protein detected across each fmol range.  \textbf(B) A log-log plot of average dNSAF vs average concentration of each group of UPS2 proteins at each fmol range: (50, 500, 5000, and 50,000 fmol).}
	\label{fig:UPS2}
\end{figure}

@@

Consequently, caution should be exercised when selecting this option even though it might provide a slight improvement in stringency (see supplemental discussion in Supporting Information).

@@

The digests were then subjected to MS/MS and the dNSAF value for each subunit along with the uNSAF value for individual isoforms were calculated by the Morpheus/MSpC pipeline (see Supplemental Methods).

@@

When the dNSAF values for all subunits for the Arabidopsis 20S proteasome including their isoforms (representing 14 distinct subunits, 10 of which exist as isoform pairs) were summed, a very close approximation of the dNSAF/actual abundance was obtained (slope=0.875) with a very strong linear correlation (R$\textsuperscript{2}$ = 0.99) over a \textasciitilde10-fold range in protein abundance.  
When each 20S proteasome subunit was analyzed individually, a strong linear response was also obtained (R$\textsuperscript{2}$ > 0.90) for a majority of subunits (Figure \ref{fig:proteasomespike} C and Supplemental Table 1).

For example, reasonably accurate concentration plots were obtained for the PAF ($\alpha$6) and PBD ($\beta$4) subunits that are encoded by the PAF1/2 and PBF1/2 gene pairs, and for the PAG ($\alpha$7) and PBF ($\beta$6) subunits that are encoded by single PAG1 and PBF1 genes (R$\textsuperscript{2}$  from 0.94 to 0.99).
Even when we calculated uNSAF values for individual isoforms added to the E. coli lysate, strong linear responses were obtained (e.g., the PAF1/PAF2 and PBD1/PBD2 pairs) with robust correlations (R$\textsuperscript{2}$ from 0.89 to 0.95) (Figure 2D).
Taken together, MSpC worked well for relative LFQ analysis of a multi-subunit complex and its individual subunits and isoforms within a complex proteomic mixture.

The Morpheus/MSpC pipeline also allowed us to calculate the respective incorporation of each paralog in the complex (see Supplemental Methods).
As shown in Figure 2E, these estimated/expected occupancies were close to unity for most subunits within both the $\alpha$ and $\beta$ rings of the 20S proteasome.
The only strong deviation was for PBD1/2 ($\beta$4), which  had a greater dNSAF value relative to other $\beta$ subunits across the experiments analyzed (Supplemental Table 1).
The calculations for uNSAF values also estimated the relative proportion of each isoform within the complex for those subunits expressed from paralogous genes.
The data obtained are similar to prior studies of the complex involving quantitative top-down proteomic analysis of purified proteasome samples using ultra violet-intrinsic fluorescence to quantify tyrosine-containing subunits \citep{russell13}.
However, our MSpC analysis provided a more complete picture as several subunit isoforms were difficult to quantify by fluorescence either because they lacked tryosine, or because their fluorescence peaks overlapped with those of other subunits/isoforms.
Notably , the protein isoform ratios measured here agree well with the expression ratios for the paralogous genes \citep{book10}, suggesting that the protein isoform abundance generally reflects the relative transcriptional activity of the gene pair.
We consistently estimated slightly more $\alpha$ ring subunits (PAA-PAG) versus $\beta$ ring subunits (PBA-PBG) in the final MSpC calculations (Figure 2E).
This deviation could represent enhanced detection of $\alpha$ ring versus $\beta$ ring subunits, or more likely that purification via the tagged $\alpha$ ring subunit PAG1 also isolated assembly intermediates comprised of only $\alpha$ ring subunits.

@@

\section {Methods}
\subsection{Sample Preparation}

@@

Protein identifications were determined using the Morpheus search engine \citep{wenger13}.  Raw data was searched with the Thermo module of Morpheus revision 151 downloaded and compiled from source code available at \url{http://sourceforge.net/projects/morpheus-ms/} using Microsoft Visual Studio 2013 professional edition.  The following parameters were used to search all databases: unknown precursor charge states - +2, +3, +4; maximum number of MS/MS peaks = 400; assign charge states - enabled; de-isotope - disabled; generate target decoy database on the fly; protease trypsin (no proline rule); maximum missed cleavages = 2; initiator methionine behavior - variable; fixed modification of carbamidomethylation of cysteine; variable modification of oxidation of methionine; maximum variable modification isoforms per peptide = 1024; precursor mass tolerance = $\pm$ 2.1 Da monoisotopic (recommended parameters to account for neutral loss); precursor monoisotopic peak correction - disabled; product mass tolerance = $\pm$ 0.025 Da monoisotopic; consider modified forms as unique peptides - false; maximum threads = 8; minimize memory usage - false.   
MSpC quantification of Universal Proteome Standard 2 (UPS2) protein sequences exploited the MS/MS analysis of individual USP2 proteins (Sigma-Aldrich) mixed at various concentrations with egg or embryo extracts from Xenopus laevis available in the PRoteomics IDEntifications (PRIDE) repository \citep{vizcaino13} using identifier - PXD000902 and the available proteomics database - pita\_v1.71.protein.name.fa.  For our analysis, both the raw MS/MS data and resulting FASTA files for the USP2 and egg and embryo proteomes were obtained from PRIDE.  The database used for searching the MS/MS data of Arabidopsis proteasomes spiked into total E. coli peptides was generated by combining Uniprot K12 E. coli reference proteome UP000000625 with a common contaminant database, and then mixing the merged dataset with FASTA sequences for all proteoforms of all known proteasome subunits and associated proteins \citep{book10} obtained from the TAIR10\_pep\_20101214 FASTA database available within The Arabidopsis Information Resource (TAIR) version 10.  All FASTA files are available for download in the Supporting Information.  The datasets were analyzed by MSpC with a 1\% PSM false discovery rate (FDR) and a 1\% protein group FDR to determine NSAF, dNSAF, and uNSAF values for each protein group. Two separate analyses were performed in which one unique, or alternatively two unique peptides were required to quantify a protein group.  NSAF values were calculated according to formulas 1a, 2a, and 3a from Figure 1 of Zhang et al. \citep{zhang10}.
	For the individual subunit analysis of the 20S proteasome the isoform incorporation rates were treated as follows. Given that each proteasome subunit should be incorporated at equal stoichiometry within the 20S particle, we then tested whether the Morpheus/MSpC pipeline could calculate the relative abundance of each subunit and the distribution of isoforms.  Here, we divided the dNSAF values for each subunit/isoform by the total number of dNSAF values for the entire complex across all eight total proteasome/E. coli lysate ratios tested.  This averaged value provided a concentration-independent ratio for the incorporation of each subunit/isoform.  We then normalized these values based on a 1/14 stoichiometry of each subunit within the complex to calculate the estimated occupancy versus the expected occupancy of each subunit.

@@

>>>>

The Morpheus MS search engine was recently designed for high-resolution, accurate-mass data obtained from Orbitrap-based instruments to provide faster matching of spectra to peptides \citep{wenger13}.
Unfortunately, no downstream automated tools are available to facilitate LFQ analysis, which can be quite challenging, if not impossible, to complete manually when accounting for shared peptides.
To overcome this bottleneck, we developed Morpheus Spectral Counter (MSpC) as the first LFQ computational tool that integrates directly with Morpheus to calculate NSAF, dNSAF, uNSAF, and corrected PSM \citep{fermin11} values in complex protein samples.
\begin{sidewaysfigure}[p]
	\centering
	\includegraphics[width=\columnwidth]{MSpC/figure1_supplemental.png}
	\mycaption{MSpC Graphic User Interface (GUI) and software flow chart}{
\textbf(A) Screenshot of the GUI.
The input requires the user to select a Morpheus search summary file containing experiments to be analyzed.  The user can optionally select a whitelist to filter output, and select an output directory.
Additional options can set peptide and protein FDR cutoffs, and method of quantification for output, including Normalized Spectral Abundance Factor (NSAF), distributed NSAF (dNSAF), and unique	NSAF (uNSAF).
A progress bar highlights completion of the analysis.
\textbf(B) Data analysis flow chart.
Experiments and groups of experiments to be analyzed are imported through the Morpheus summary file.
PSMs and protein groups are filtered at the specified FDR cutoff with a default of 1\%.}
	\label{fig:GUIflow}
\end{sidewaysfigure}
MSpC is fully automated, and only requires a Morpheus search summary file (summary.tsv) as input.
The user interface (see Figure \ref{fig:GUIflow} (A)) allows one to select the summary file and displays the raw MS/MS files that will be analyzed by MSpC.
Due to shared peptides being attributed to only one instance of a protein group in Morpheus's PSM file, PSMs are re-matched to all possible protein groups.
PSMs are then cataloged as shared or as unique (distinctly matching one protein group) to generate NSAF, dNSAF, and uNSAF outputs.
Finally, the output can be filtered for proteins of interest by specifying a comma delimited file containing unique identifiers and descriptions.
Some important features of MSpC are its ability to handle fractionation experiments as input, and the ability to whitelist proteins of interest in the output by specifying a csv file (see Tutorial \ref{sec:tutorial}).
Options exist to specify global PSM and protein group FDR rates (thus avoiding increased FDRs when one analyzes many experiments at once), to output NSAF, dNSAF, and uNSAF values, to require a minimum number of unique peptides to quantify a protein, and to specify an output directory.
A progress bar indicates completion of the analysis by MSpC.
To validate the accuracy of MSpC, we analyzed two MS/MS datasets available in PRIDE that were previously generated by high-energy collision-induced dissociation using Thermo Q-Exactive Orbitrap instruments.
\begin{figure}[p]
	\centering
	\includegraphics[width=\columnwidth]{MSpC/figure1.png}
	\mycaption{Confirmation of MSpC accuracy by analysis of MS/MS datasets generated with the Universal Proteome Standard 2 (UPS2)}{The array of UPS2 standards were spiked into Xenopus laevis egg \textbf(Top) and embryo \textbf(Bottom) extracts at a range of concentrations.  Following MS/MS analysis, dNSAF values for each protein were determined by Morpheus and MSpC.  \textbf(A) A log-log plot of dNSAF versus concentration for each UPS2 protein detected across each fmol range.  \textbf(B) A log-log plot of average dNSAF vs average concentration of each group of UPS2 proteins at each fmol range: (50, 500, 5000, and 50,000 fmol).}
	\label{fig:UPS2}
\end{figure}

@@

Consequently, caution should be exercised when selecting this option even though it might provide a slight improvement in stringency (see Discussion on Requiring Two Peptide \ref{sec:discussiontwopep}).

@@

The digests were then subjected to MS/MS and the dNSAF value for each subunit along with the uNSAF value for individual isoforms were calculated by the Morpheus/MSpC pipeline (see Methods \ref{sec:methods}).

@@

When the dNSAF values for all subunits for the Arabidopsis 20S proteasome including their isoforms (representing 14 distinct subunits, 10 of which exist as isoform pairs) were summed, a very close approximation of the dNSAF/actual abundance was obtained (slope=0.875) with a very strong linear correlation (R$\textsuperscript{2}$ = 0.99) over a \textasciitilde10-fold range in protein abundance.  
When each 20S proteasome subunit was analyzed individually, a strong linear response was also obtained (R$\textsuperscript{2}$ > 0.90) for a majority of subunits (Figure \ref{fig:proteasomespike} C and Table \ref{table:MSpC}).

For example, reasonably accurate concentration plots were obtained for the PAF ($\alpha$6) and PBD ($\beta$4) subunits that are encoded by the PAF1/2 and PBF1/2 gene pairs, and for the PAG ($\alpha$7) and PBF ($\beta$6) subunits that are encoded by single PAG1 and PBF1 genes (R$\textsuperscript{2}$  from 0.94 to 0.99).
Even when we calculated uNSAF values for individual isoforms added to the E. coli lysate, strong linear responses were obtained (e.g., the PAF1/PAF2 and PBD1/PBD2 pairs) with robust correlations (R$\textsuperscript{2}$ from 0.89 to 0.95) (Figure {fig:proteasomespike} D).
Taken together, MSpC worked well for relative LFQ analysis of a multi-subunit complex and its individual subunits and isoforms within a complex proteomic mixture.

The Morpheus/MSpC pipeline also allowed us to calculate the respective incorporation of each paralog in the complex (see Isoform Incorporation Methods \ref{subsec:isoform}).
As shown in Figure \ref{fig:proteasomespike}E, these estimated/expected occupancies were close to unity for most subunits within both the $\alpha$ and $\beta$ rings of the 20S proteasome.
The only strong deviation was for PBD1/2 ($\beta$4), which  had a greater dNSAF value relative to other $\beta$ subunits across the experiments analyzed (see Table \ref{table:MSpC}).
The calculations for uNSAF values also estimated the relative proportion of each isoform within the complex for those subunits expressed from paralogous genes.
The data obtained are similar to prior studies of the complex involving quantitative top-down proteomic analysis of purified proteasome samples using ultra violet-intrinsic fluorescence to quantify tyrosine-containing subunits \citep{russell13}.
However, our MSpC analysis provided a more complete picture as several subunit isoforms were difficult to quantify by fluorescence either because they lacked tryosine, or because their fluorescence peaks overlapped with those of other subunits/isoforms.
Notably , the protein isoform ratios measured here agree well with the expression ratios for the paralogous genes \citep{book10}, suggesting that the protein isoform abundance generally reflects the relative transcriptional activity of the gene pair.
We consistently estimated slightly more $\alpha$ ring subunits (PAA-PAG) versus $\beta$ ring subunits (PBA-PBG) in the final MSpC calculations (Figure \ref{fig:proteasomespike} E).
This deviation could represent enhanced detection of $\alpha$ ring versus $\beta$ ring subunits, or more likely that purification via the tagged $\alpha$ ring subunit PAG1 also isolated assembly intermediates comprised of only $\alpha$ ring subunits.

@@

\section {Methods}
\label{sec:methods}
\subsection{Sample Preparation}

@@

Protein identifications were determined using the Morpheus search engine \citep{wenger13}.  Raw data was searched with the Thermo module of Morpheus revision 151 downloaded and compiled from source code available at \url{http://sourceforge.net/projects/morpheus-ms/} using Microsoft Visual Studio 2013 professional edition.  The following parameters were used to search all databases: unknown precursor charge states - +2, +3, +4; maximum number of MS/MS peaks = 400; assign charge states - enabled; de-isotope - disabled; generate target decoy database on the fly; protease trypsin (no proline rule); maximum missed cleavages = 2; initiator methionine behavior - variable; fixed modification of carbamidomethylation of cysteine; variable modification of oxidation of methionine; maximum variable modification isoforms per peptide = 1024; precursor mass tolerance = $\pm$ 2.1 Da monoisotopic (recommended parameters to account for neutral loss); precursor monoisotopic peak correction - disabled; product mass tolerance = $\pm$ 0.025 Da monoisotopic; consider modified forms as unique peptides - false; maximum threads = 8; minimize memory usage - false.   
MSpC quantification of Universal Proteome Standard 2 (UPS2) protein sequences exploited the MS/MS analysis of individual USP2 proteins (Sigma-Aldrich) mixed at various concentrations with egg or embryo extracts from Xenopus laevis available in the PRoteomics IDEntifications (PRIDE) repository \citep{vizcaino13} using identifier - PXD000902 and the available proteomics database - pita\_v1.71.protein.name.fa.  For our analysis, both the raw MS/MS data and resulting FASTA files for the USP2 and egg and embryo proteomes were obtained from PRIDE.  The database used for searching the MS/MS data of Arabidopsis proteasomes spiked into total E. coli peptides was generated by combining Uniprot K12 E. coli reference proteome UP000000625 with a common contaminant database, and then mixing the merged dataset with FASTA sequences for all proteoforms of all known proteasome subunits and associated proteins \citep{book10} obtained from the TAIR10\_pep\_20101214 FASTA database available within The Arabidopsis Information Resource (TAIR) version 10.  All FASTA files are available for download in the Supporting Information.  The datasets were analyzed by MSpC with a 1\% PSM false discovery rate (FDR) and a 1\% protein group FDR to determine NSAF, dNSAF, and uNSAF values for each protein group. Two separate analyses were performed in which one unique, or alternatively two unique peptides were required to quantify a protein group.  NSAF values were calculated according to formulas 1a, 2a, and 3a from Figure 1 of Zhang et al. \citep{zhang10}.

\subsection {Isoform Incorporation Rates}
\label{subsec:isoform}
	For the individual subunit analysis of the 20S proteasome the isoform incorporation rates were treated as follows. Given that each proteasome subunit should be incorporated at equal stoichiometry within the 20S particle, we then tested whether the Morpheus/MSpC pipeline could calculate the relative abundance of each subunit and the distribution of isoforms.  Here, we divided the dNSAF values for each subunit/isoform by the total number of dNSAF values for the entire complex across all eight total proteasome/E. coli lysate ratios tested.  This averaged value provided a concentration-independent ratio for the incorporation of each subunit/isoform.  We then normalized these values based on a 1/14 stoichiometry of each subunit within the complex to calculate the estimated occupancy versus the expected occupancy of each subunit.

@@

\section {Discussion on Requiring Two Unique Peptides to Quantify a Protein}
\label{sec:discussiontwopep}
Occasionally, some researchers may want to use a more stringent criterion for quantification such as requiring a protein to have more than one unique identifying peptide. To see how this might affect our data analysis, we re-analyzed our results shown in Figure \ref{fig:UPS2twopep}, this time requiring two unique peptides to quantify an individual protein. The results point to a very small increase in linearity observed in the average plots; however, there is a slight decrease in linearity for the egg sample (0.886 to 0.865) and a larger decrease in linearity for the individual UPS2 protein plot for the embryo sample (0.827 to 0.723). The decrease in linearity in the embryo sample is due to the analysis removing a low abundance UPS2 protein (O00762ups) identified with only one unique peptide. While some have suggested that requiring more than one unique peptide to identify a protein is an ideal approach, requiring two peptides for identifications in database searches reduces the number of protein identifications in the target database more than those in the decoy database and results in increased false discovery rates . While we recognize that researchers may want to implement more stringent requirements than what is typically used in database searching to quantify a set of proteins, there are two cases where requiring two unique peptides may not be ideal in a quantitative analysis. Firstly, low abundance proteins that have few PSMs might be identified by only a single peptide and thus be erroneously thrown out of the analysis. Secondly, there may be only one unique peptide that can differentiate between families of homologous proteins. In this second case, requiring two unique peptides would remove these homologous proteins from the MSpC analysis, even if they had a large number of PSMs. Because of these reasons and because of the decreased linearity observed when requiring proteins to have two unique peptides (Figure \ref{fig:UPS2twopep}) as compared to one unique peptide (Figure \ref{fig:UPS2}), we suggest caution in requiring more than one unique peptide per protein.

\section {Tutorial}
\label{sec:tutorial}

\includepdf[pages=-]{MSpC/Tutorial.pdf}
